---
title: "5.1. Install Packages"
author: Adapted from Software Carpentry; modified by Maina,Stephanie
date: "13th of March 2018"
output: html_document
---
  
  ```{r echo=FALSE}
library(knitr)
library(rmarkdown)
```

```{r setup, echo=FALSE}
#opts_chunk$set(echo = FALSE, cache=FALSE)
read_chunk('/Users/stephdagata/Documents/GitHub/MQE-R-WORKSHOP-2018/Scripts/Day 1/Day1_5.1.Install_Packages.R')
```

## Installing and managing packages

**LISTS OF PACKAGES**
```{r eval=F}
browseURL("http://cran.r-project.org/web/views/")  # Opens CRAN Task Views in browser
```
Opens "Available CRAN Packages By Name" (from UCLA mirror) in browser
```{r eval=F}
browseURL("http://cran.stat.ucla.edu/web/packages/available_packages_by_name.html")
```
See also CRANtastic
```{r eval=F}
browseURL("http://crantastic.org/")
```

+ **See current packages**

```{r eval=F}
library()  # Brings up editor list of installed packages
search()   # Shows packages that are currently loaded
```

## To install and use packages

Can use menus: 
  + Tools > Install Packages... (or use Package window)
  + Or can use scripts, which can be saved in incorporated in source

```{r eval=F}
install.packages("ggplot2")  # Downloads package from CRAN and installs in R
?install.packages
```

```{r eval=F}
library("ggplot2")  # Make package available; often used for loading in scripts
require("ggplot2")  # Preferred for loading in functions; maybe better?
library(help = "ggplot2")  # Brings up documentation in editor window
```

+ **UPDATE PACKAGES**
In RStudio, Tools > Check for Package Updates

```{r eval=F}
update.packages()  # Checks for updates; do periodically
?update.packages
```

+ **UNLOAD/REMOVE PACKAGES**

By default, all loaded packages are unloaded when R quits
Can also open Packages window and manually uncheck or can use this code to unload packages
```{r eval=F}
detach("package:ggplot2", unload = TRUE)
?detach
```

+ **To permanently remove (delete) package**

```{r eval=F}
install.packages("psytabs")  # Adds psytabs
remove.packages("psytabs")   # Deletes it
?remove.packages
```

## Getting help with R
This is a summary of the information you can 
find here https://www.r-project.org/help.html

#### R Help: 'help()' and '?'
The 'help()' function and '?' help operator in R provide access to the documentation pages for **R functions, data sets, 
and other objects**, both for packages in the standard R distribution and for contributed packages. 
For example, to access documentation for the standard mean function, 
for example, enter the command:'help(mean)' or 'help("mean")', or '?mean' or '?"mean"' (the quotes are optional).

You may also use the 'help()' function to access information about a **package** in your library - 
for example:

```{r eval=F}
help(package="MASS")
```

which displays an index of available help pages for the package along with some other information.

Help pages for functions usually include a section with executable examples illustrating how the functions work. 
You can execute these examples in the current R session via the 'example()' 

```{r eval=F}
example(mean)
```

#### Searching for Help Within R
The 'help()'  function and '?' operator are useful only if you already know the name of the function that you wish to use. 
There are also facilities in the standard R distribution for discovering functions and other objects. 
The following functions cast a progressively wider net. Use the help system to obtain complete documentation for these functions: for example, `apropos()` or `?apropos`.

The `apropos()` function searches for objects, 
including functions, directly accessible in the current R session that have names that include a specified character string. 
This may be a literal string or a regular expression to be used for pattern-matching. 
By default, string matching by 'apropos()' is case-insensitive. For example, 

```{r eval=F}
apropos("^glm") 
```

returns the names of all accessible objects that start with the (case-insensitive) characters "glm"

#### R Help on the internet
There are internet search sites that are specialized for R searches
including 'search.r-project.org' (which is the site used by RSiteSearch) and `Rseek.org`.

It is also possible to use a general search site like **Google**,
by qualifying the search with "R" or the name of an R package (or both). 
**It can be particularly helpful to paste an error message into a search engine** to find out whether others have solved a problem that you encountered.

