---
title: "4.4. Making Choices"
author: Adapted from Software Carpentry; modified by Maina,Stephanie
date: "13th of March 2018"
output: html_document
---

```{r echo=FALSE}
library(knitr)
library(rmarkdown)
```

```{r setup, echo=FALSE}
#opts_chunk$set(echo = FALSE, cache=FALSE)
read_chunk('/Users/stephdagata/Documents/GitHub/MQE-R-WORKSHOP-2018/Scripts/Day 1/Day1_4.4.Making Choices.R')
```

**Questions**

  + How do I make choices using `if` and `else` statements?
  + How do I compare values?
  + How do I save my plots to a PDF file?

**Objectives**

  + Save plot(s) in a PDF file.
  + Write conditional statements with `if` and `else`.
  + Correctly evaluate expressions containing `&` ('and') and `|` ('or').

Our previous lessons have shown us how to manipulate data, define our own functions, and repeat things. However, the programs we have written so far always do the same things, regardless of what data they're given. We want programs to make choices based on the values they are manipulating.
```

The function `pdf` redirects all the plots generated by R into a pdf file, which in this case we have named "inflammation-01.pdf". After we are done generating the plots to be saved in the pdf file, we stop R from redirecting plots with the function `dev.off`.

### Overwriting Plots

If you run pdf multiple times without running `dev.off`, you will save plots to the most recently opened file. However, you won't be able to open the previous pdf files because the connections were not closed. In order to get out of this situation, you'll need to run `dev.off` until all the pdf connections are closed. You can check your current status using the function `dev.cur`. If it says "pdf", all your plots are being saved in the last pdf specified. If it says "null device" or "RStudioGD", the plots will be visualized normally.

We can update the `analyze` function so that it always saves the plots in a pdf. But that would make it more difficult to interactively test out new changes. It would be ideal if `analyze` would either save or not save the plots based on its input.

## Conditionals

In order to update our function to decide between saving or not, we need to write code that automatically decides between multiple options. The tool R gives us for doing this is called a conditional statement, and looks like this:

```{r MC1}
#num <- 37
#if (num > 100) {
#  print("greater")
#} else {
#  print("not greater")
#}
#print("done")
```

The second line of this code uses an `if` statement to tell R that we want to make a choice. If the following test is true, the body of the `if` (i.e., the lines in the curly braces underneath it) are executed. If the test is false, the body of the `else` is executed instead. Only one or the other is ever executed:

```{r MC2}
#num > 100
```

And as you likely guessed, the opposite of `FALSE` is `TRUE`.

```{r MC3}
#num < 100
```
Conditional statements don't have to include an `else`. If there isn't one, R simply does nothing if the test is false:

```{r MC4}
#num <- 53
#if (num > 100) {
#  print("num is greater than 100")
#}
```
We can also chain several tests together when there are more than two options. This makes it simple to write a function that returns the sign of a number:

```{r MC5}
#sign <- function(num) {
#  if (num > 0) {
#    return(1)
#  } else if (num == 0) {
#    return(0)
#  } else {
#    return(-1)
#  }
#}
```

```{r MC6}
#sign(-3)
```

```{r MC7}
#sign(0)
```

```{r MC8}
#sign(2/3)
```
Note that when combining `else` and `if` in an `else if` statement, the `if` portion still requires a direct input condition. This is never the case for the else statement alone, which is only executed if all other conditions go unsatisfied. Note that the test for equality uses two equal signs, `==`.

### Other Comparisons

Other tests include greater than or equal to (`>=`), less than or equal to (`<=`), and not equal to (`!=`).

We can also combine tests. An ampersand, `&`, symbolizes "and". A vertical bar, `|`, symbolizes "or". `&` is only true if both parts are true:

```{r MC9}
#if (1 > 0 & -1 > 0) {
#    print("both parts are true")
#} else {
#  print("at least one part is not true")
#}
```
while `|` is true if either part is true:

```{r MC10}
#if (1 > 0 | -1 > 0) {
#    print("at least one part is true")
#} else {
#  print("neither part is true")
#}
```

In this case, "either" means "either or both", not "either one or the other but not both".

### Choosing Plots Based on Data

Write a function `plot_dist` that plots a boxplot if the length of the vector is greater than a specified threshold and a stripchart otherwise. To do this you'll use the R functions `boxplot` and `stripchart`.

```{r MC11,echo=FALSE}
#plot_dist <- function(x, threshold) {
#  if (length(x) > threshold) {
#    boxplot(x)
#  } else {
#    stripchart(x)
#  }
#}
```

```{r MC11a,echo=-1}
#dat <- read.csv("data/inflammation-01.csv", header = FALSE)
#plot_dist(dat[, 10], threshold = 10)     # day (column) 10
```

```{r MC12}
#plot_dist(dat[1:5, 10], threshold = 10)  # samples (rows) 1-5 on day (column) 10
```

### Solution:
```{r MC13,eval=FALSE}
#plot_dist <- function(x, threshold) {
#  if (length(x) > threshold) {
#    boxplot(x)
#  } else {
#    stripchart(x)
#  }
#}
```

### Histograms Instead

One of your collaborators prefers to see the distributions of the larger vectors as a histogram instead of as a boxplot. In order to choose between a histogram and a boxplot we will edit the function `plot_dist` and add an additional argument `use_boxplot`. By default we will set `use_boxplot` to `TRUE` which will create a boxplot when the vector is longer than `threshold`. When `use_boxplot` is set to `FALSE`, `plot_dist` will instead plot a histogram for the larger vectors. As before, if the length of the vector is shorter than `threshold`, `plot_dist` will create a stripchart. A histogram is made with the `hist` command in R.

```{r MC14,echo=-1}
#plot_dist <- function(x, threshold, use_boxplot = TRUE) {
#   if (length(x) > threshold & use_boxplot) {
#       boxplot(x)
#   } else if (length(x) > threshold & !use_boxplot) {
#       hist(x)
#   } else {
#       stripchart(x)
#   }
#}
```

```{r MC16}
#dat <- read.csv("data/inflammation-01.csv", header = FALSE)
#plot_dist(dat[, 10], threshold = 10, use_boxplot = TRUE)   # day (column) 10 - create boxplot
```

```{r MC17}
#plot_dist(dat[, 10], threshold = 10, use_boxplot = FALSE)  # day (column) 10 - create histogram
```
```{r MC18}
#plot_dist(dat[1:5, 10], threshold = 10)                    # samples (rows) 1-5 on day (column) 10
```

```{r MC19,eval=FALSE}
#plot_dist <- function(x, threshold, use_boxplot = TRUE) {
#   if (length(x) > threshold & use_boxplot) {
#       boxplot(x)
#   } else if (length(x) > threshold & !use_boxplot) {
#       hist(x)
#   } else {
#      stripchart(x)
#   }
#}
```

### Find the Maximum Inflammation Score

Find the file containing the patient with the highest average inflammation score. Print the file name, the patient number (row number) and the value of the maximum average inflammation score.

Tips:

  + 1. Use variables to store the maximum average and update it as you go through files and patients.
  + 2.You can use nested loops (one loop is inside the other) to go through the files as well as through the patients in each file (every row).
  
Complete the code below:

```{r MC19a}
#filenames <- list.files(path = "data", pattern = "inflammation.*csv", full.names = TRUE)
#filename_max <- "" # filename where the maximum average inflammation patient is found
#patient_max <- 0 # index (row number) for this patient in this file
#average_inf_max <- 0 # value of the average inflammation score for this patient
#for (f in filenames) {
#  dat <- read.csv(file = f, header = FALSE)
#  dat.means = apply(dat, 1, mean)
#  for (patient_index in length(dat.means)){
#    patient_average_inf = dat.means[patient_index]
#    # Add your code here ...
#  }
#}
```

```{r MC20}
#print(filename_max)
```

```{r MC21}
#print(patient_max)
```

```{r MC22}
#print(average_inf_max)
```
### Solution

```{r MC23,eval=FALSE}
# Add your code here ...
#if (patient_average_inf > average_inf_max) {
#  average_inf_max = patient_average_inf
#  filename_max <- f
#  patient_max <- patient_index
#}
```

## Saving Automatically Generated Figures
Now that we know how to have R make decisions based on input values, let's update `analyze`:

```{r MC24}
#analyze <- function(filename, output = NULL) {
#  # Plots the average, min, and max inflammation over time.
#  # Input:
#  #    filename: character string of a csv file
#  #    output: character string of pdf file for saving
#  if (!is.null(output)) {
#    pdf(output)
#  }
#  dat <- read.csv(file = filename, header = FALSE)
#  avg_day_inflammation <- apply(dat, 2, mean)
#  plot(avg_day_inflammation)
#  max_day_inflammation <- apply(dat, 2, max)
#  plot(max_day_inflammation)
#  min_day_inflammation <- apply(dat, 2, min)
#  plot(min_day_inflammation)
#  if (!is.null(output)) {
#    dev.off()
#  }
#}
```

We added an argument, `output`, that by default is set to `NULL`. An `if` statement at the beginning checks the argument `output` to decide whether or not to save the plots to a pdf. Let's break it down. The function `is.null` returns `TRUE` if a variable is `NULL` and `FALSE` otherwise. The exclamation mark, `!`, stands for "not". Therefore the line in the `if` block is only executed if `output` is "not null".

```{r MC25}
#output <- NULL
```

```{r MC26}
#is.null(output)
```

```{r MC27}
#!is.null(output)
```

Now we can use `analye` interactively, as before,

```{r MC28,eval=F}
#analyze("data/inflammation-01.csv")
```
but also use it to save plots,

```{r MC29,eval=F}
#analyze("data/inflammation-01.csv", output = "inflammation-01.pdf")
```

Before going further, we will create a directory `results` for saving our plots. It is good practice in data analysis projects to save all output to a directory separate from the data and analysis code. You can create this directory using the shell command mkdir, or the R function `dir.create()`

```{r MC30}
#dir.create("results")
```
Now run `analyze` and save the plot in the `results` directory,

```{r MC31,eval=F}
#analyze("data/inflammation-01.csv", output = "results/inflammation-01.pdf")
```

This now works well when we want to process one data file at a time, but how can we specify the output file in `analyze_all`? We need to do two things:

  + Substitute the filename ending "csv" with "pdf".
  + Save the plot to the `results` directory.
  
To change the extension to "pdf", we will use the function `sub`,

```{r MC32}
#f <- "inflammation-01.csv"
#sub("csv", "pdf", f)
```

To add the "data" directory to the filename use the function `file.path`,

```{r MC33}
#file.path("results", sub("csv", "pdf", f))
```
Now let's update `analyze_all`:

```{r MC34}
#analyze_all <- function(pattern) {
#  # Directory name containing the data
#  data_dir <- "data"
#  # Directory name for results
#  results_dir <- "results"
#  # Runs the function analyze for each file in the current working directory
#  # that contains the given pattern.
#  filenames <- list.files(path = data_dir, pattern = pattern)
#  for (f in filenames) {
#    pdf_name <- file.path(results_dir, sub("csv", "pdf", f))
#    analyze(file.path(data_dir, f), output = pdf_name)
#  }
#}
```

Now we can save all of the results with just one line of code:

```{r MC35}
#analyze_all("inflammation.*csv")
```

Now if we need to make any changes to our analysis, we can edit the analyze function and quickly regenerate all the figures with `analyze_all`.

### Changing the Behavior of the Plot Command

One of your collaborators asks if you can recreate the figures with lines instead of points. Find the relevant argument to `plot` by reading the documentation (`?plot`), update `analyze`, and then recreate all the figures with `analyze_all`.

```{r MC36,eval=FALSE}
#analyze <- function(filename, output = NULL) {
#  # Plots the average, min, and max inflammation over time.
#  # Input:
#  #    filename: character string of a csv file
#  #    output: character string of pdf file for saving
#  if (!is.null(output)) {
#    pdf(output)
#  }
#  dat <- read.csv(file = filename, header = FALSE)
#  avg_day_inflammation <- apply(dat, 2, mean)
#  plot(avg_day_inflammation, type = "l")
#  max_day_inflammation <- apply(dat, 2, max)
#  plot(max_day_inflammation, type = "l")
#  min_day_inflammation <- apply(dat, 2, min)
#  plot(min_day_inflammation, type = "l")
#  if (!is.null(output)) {
#    dev.off()
#  }
#}
```

## Key Points

  + Save a plot in a pdf file using `pdf("name.pdf")` and stop writing to the pdf file with `dev.off()`
  + Use `if (condition)` to start a conditional statement, `else if (condition)` to provide additional tests, and `else` to provide a default.
  + The bodies of conditional statements must be surrounded by curly braces `{ }`.
  + Use `==` to test for equality.
  + `X & Y` is only true if both X and Y are true.
  + `X | Y` is true if either X or Y, or both, are true.
