---
title: "4.5.Missing Data"
author: "Modified from nice R code"
date: "13 of March 2018"
output: html_document
---

```{r echo=FALSE}
library(knitr)
library(rmarkdown)
```

```{r setup, echo=FALSE}
#opts_chunk$set(echo = FALSE, cache=FALSE)
read_chunk('/Users/stephdagata/Documents/GitHub/MQE-R-WORKSHOP-2018/Scripts/Day 1/Day1_4.5.Missing Data.R')
```

Missing data are a fact of life in biology. Individuals die, equipment breaks, you forget to measure something, you can't read your writing, etc.

If you load in data with blank cells, they will appear as an `NA` value.

```{r MD1,echo=-1}
data <- read.csv("data/seed_root_herbivores.csv")
```

Some data to play with.

```{r MD2}
x <- data$Height[1:10]
```

If the 5th element was missing

```{r MD3}
x[5] <- NA
```

This is what it would look like:

Note that this is not a string "NA"; that is something different entirely.

Treat a missing value as a number that could stand in for anything. So what is

```{r MD4}
1 + NA
```

```{r MD5}
1 * NA
```

```{r MD6}
NA + NA
```

These are all NAs because if the input could be anything, the output could be anything.

What is the value of this:

```{r MD7}
mean(x)
```

It's `NA` too because `x[1] + x[2] + NA + ...` must be `NA`. And then `NA/length(x)` is also `NA`.

This is a pretty common situation for data, so the mean function takes an `na.rm` argument

```{r MD8}
mean(x, na.rm = TRUE)
```

`sum` takes the same argument too:

```{r MD9}
sum(x, na.rm = TRUE)
```

Be careful though:

```{r MD10}
sum(x, na.rm = TRUE)/length(x)  # not the mean!
```

```{r MD11}
mean(x, na.rm = TRUE)
```

The `na.omit`function will strip out all NA values:

```{r MD12}
na.omit(x)
```


So we can do this:

```{r MD13}
length(na.omit(x))
```

You can't test for `NA`-ness with `==`:

```{r MD14}
x == NA
```

(why not?)

Use `is.na` instead:

```{r MD15}
is.na(x)
```

So `na.omit` is (roughly) equivalent to

```{r MD16}
x[!is.na(x)]
```


## Exercise

Our standard error function doesn't deal well with missing values:

```{r MD17}
standard.error <- function(x) {
    v <- var(x)
    n <- length(x)
    sqrt(v/n)
}
```

Can you write one that always filters missing values?

If we get time, we'll talk about how to write one that optionally gets rid of missing values.

## Other special values:

Positive and negative infinities

```{r MD18}
1/0
```

```{r MD19}
-1/0
```

Not A Number (different to `NA`, but usually treatable the same way).


We saw `NULL` before. It's the weirdest.

```{r MD20}
0/0
```
