---
title: "Day 1: R Data Manipulation - Getting started"
#author: "XXX"
date: "27 October 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Using R as a calculator

### 1.1. Doing arithmetic with R

The simplest thing you could do with R is to arithmetic

```{r arithmetic}
1+100
```
Here, we've added 1 and and 100 together to make 101. The [1] preceeding this we will explain in a minute. For now, think of it as something that indicates output.

Order of operations is same as in maths class (from highest to lowest precedence)

  + Brackets
  + Exponents
  + Divide
  + Multiply
  + Add
  + Subtract

What will this evaluate to?

```{r}
3+5*2
```

The "caret" symbol (or "hat") is the exponent (to-the-power-of) operator (read `x ^ y` as "x to the power of y"). What will this evaluate to?

```{r}
3+5*2^2
```

Use brackets (actually parentheses) to group to force the order of evaluation if it differs from the default, or to set your own order.

```{r}
(3+5)*2
```
But this can get unweidly when not needed:

```{r}
(3+(5*(2^2))) # hard to read
3+5*2^2 #easier to read, once you know some rules
3+5*(2^2) # if you forget some rules, this might help
```
See `?Arithmetic` for more information, and two more operators (you can also get there by `?"+"` (note the quotes)

If R thinks that the statement is incomplete, it will change the prompt from `>` to `+` indicating that it is expecting more input. This is not an addition sign! 
Try: 


```{r error=TRUE}
3+5*(2^2
```

Press `"Esc"` if you want to cancel this statement and return to the prompt.

### 1.2. Comparing objects with R
The usual sort of comparison operators are available:

```{r}
1 == 1 #equality (note two equals signes, read as "is equal to")
```

```{r}
1 == 2
```

```{r}
1 != 2 #inequality (read as "is not equal to")
```

```{r}
1<2 # less than
```

```{r}
1<=1 # less than or equal to"
```

```{r}
1>0 # greater than
```

```{r}
1>= -9 # greater than or equal to
```
See `?Comparison` for more information (you can also get there by `help("==")`.

### 1.3. Scientific notation with R

Really small numbers get a scientific notation:

```{r}
2/10000
```
which you can write in too:

```{r}
2e-04
```
Read `e-XX` as "multiplied by `10^XX"`, so `2e-4` is `2 * 10^(-4)`.

### 1.4. Mathematical functions with R

R has many built in mathematical functions that will work as you would expect:

```{r}
sin(1) # trigonometric functions
```

```{r}
asin(1) # inverse sin (also for cos and tan)"
```

```{r}
log(1) #natural logarithm
```

```{r}
log10(100) # base-10 logarithm  
```

```{r}
log2(100) # base-2 logarithm  
```

```{r}
exp(0.5) #e^(1/2)
```

Plus things like probability density functions for many common distributions, and other mathematical functions (e.g., Gamma, Beta, Bessel). If you need it, it's probably there

#2. Data objects in R

### 2.1. Variables and assignment
You can assign values to variables using the assignment operator `<-`, like this:

```{r}
x <-1/40
```
and now the variable `x` contains the **value** `0.025`

```{r}
x
```

(note that it does not contain the fraction 1/40, it contains a decimal approximation of this fraction. This appears exact in this case, but it is not. These decimal approximations are called "floating point numbers" and at some point you will probably end up having to learn more about them than you'd like).

Look up at the top right pane of RStudio, and you'll see that this has appeared in the "Workspace" pane.

Our variable `x` can be used in place of a number in any calculation that expects a number.

```{r}
log(x)
```

```{r}
sin(x)
```
The right hand side of the assignment can be any valid R expression.

It is also possible to use the `=` operator for assignment:

```{r}
x=1/40
```
but this is much less common among R users. The most important thing is to be **consistent** with the operator you use. There are occasionally places where it is less confusing to use `<-` than `=`, and it is the most common symbol used in the community. So I'd recommend `<-`.

Notice that assignment does not print a value.

```{r}
x<-100
```
Notice also that variables can be reassigned.
`x`used to contain the value 0.025 and and now it has the value 100).

```{r}
x
```


Assignment values can contain the variable being assigned to: What will `x` contain after running this?

```{r}
x<-x+1
x
```

###2.2. Vectors

R was designed for people who do data analysis. There is a reason why "data" is a more common term than "datum" - generally you have more than one piece of data (although the Guardian argues that this distinction is old fashioned). As a result in R all data types are actually vectors. 
So the number '1' is actually a vector of numbers that happens to be of length 1.

```{r}
1
length(1) # length(): get the length of a vector
```

### Build a specific vector
To build a vector, use the `c()` function (`c` stands for "concatenate")
 
 
```{r}
x<- c(1,2,40,1234)
x
```

```{r}
length(x)
```
(notice how RStudio has updated its description of x. If you click it, you'll get an option to alter it, which is rarely what you want to do).

This is a deep piece of engineering in the design; most of R thinks quite happily in terms of vectors. If you wanted to double all the values in the vector, just multiply it by 2:

```{r}
2*x
```
You can get the maximum value:
```{r}
max(x)
```

... the minimum value 

```{r}
min(x)
```
... the sum

```{r}
sum(x)
```
... the mean value

```{r}
mean(x)
```

and so on. There are huge numbers of functions that operate on vectors. It is more common that functions will than that they won't.

Vectors can be summed together:

```{r}
y<- c(0.1,0.2,0.3,0.4)
x+y
```

and they can be concatenate together:

```{r}
c(x,y)
```

... and scalars can be added to them

```{r}
x+0.1
```

**Be careful** though: if you add/multiply together vectors that are of different lengths, but the lengths factor, R will silently "recycle" the length of the shorter one:

```{r}
x
```

```{r}
x*c(-2,2)
```
The **first** and **third** element have been multiplied by **-2** while the **second** and **fourth** element are multiplied by *2*.

If the length of the shorter vector is not a factor of the length of the longer vector, you will get a warning, but **the calculation will happen anyway**:

```{r}
x*c(-2,0,2)
```
This is almost never what you want. Pay attention to warnings. Note that Warnings are different to Errors. We just saw a warning, where what happened is (probably) undesirable but not fatal. You'll get Errors where what happened has been deemed unrecoverable. For example:

```{r error=TRUE}
x+z
```
Just as with the scalars, we can do comparisons. This returns a new vector of TRUE and FALSE indicating which elements are less than 10:

```{r}
x<10
```
You can do a vector-vector comparison too:

```{r }
x<y
```
And combined arithmetic operations with comparison operations. Both sides of the expression are fully evaluated before the comparison takes place:

```{r error=TRUE}
x>1/Y
```
Be careful with comparisons: This compares the first element with -20, the second with 20, the third with -20 and the fourth with 20.

```{r}
x>=c(-20,20)
```

This does nothing sensible, really, and warns you again:

```{r}
x == c(-2,0,2)
```

All the comparison operators work in fairly predictible ways:

```{r}
x == 40
```
```{r}
x != 2
```


#### Define a vector with a sequence

Sequences are easy to make, and often useful. Integer sequences can be made with the colon operator:

```{r}
3:10 # sequence 3,4,5,...,10
```
Which also works backwards:

```{r}
10:3
```

Step in different sizes with `seq()`

```{r}
seq(3,10,by=2)
```
```{r}
seq(3,10,length=10)
```

Now we will see the meaning of the `[1]` term - this indicates that you're looking at the first element of a vector. If you make a really long vector, you'll see new numbers:

```{r}
seq(3,by=2,length=100)
```

#### Finding specific elements in a vector

First, assign an element to x

```{r}
x=seq(3,by=2,length=100) # x is assigned a vector of length 100, starting from 3, with an increment of 2
x
```

Find the 1st element of x using `[]`

```{r}
x[1] # 1st element
```
```{r}
x[2] # 2nd element
```

```{r}
x[100] # 100th element
```


###2.3. Matrix

Define a matrix (assign to y) with 9 rows and 3 columns using the data in x

**Note**: x has been used several time previously. If you wish to assign a new value to x, clean x first by removing it from the workspace.

```{r error=TRUE}
x  
rm(x)
x # if an error meesage "object x not found", x has been removed succesfully
```

```{r}
x=1:27
y <- matrix(data=x,nrow=9,ncol=3)
y
```

Dimension of y with `dim()`
```{r}
dim(y)
```

Result provides a vector of 2. The first element is the number of lines (here 9) and the second element is the number of columns (here 3 columns).

Extract a value from y with `[]`

```{r}
y[1,2] # The element at the crossroad between the line 1 and column 2
```

Extract a line from y with `[]`

```{r}
y[1,] # line 1
```
Extract a column from y with `[]`

```{r}
y[,1] # column 1
```

#### Extract a vector from y using the numerical indices in x
```{r}
y[x]  # referencing an object with another object

```

#### Define a 3-dimensional array (assign to z) with 3 rows, 3 columns and 3 layers, using the data in x

```{r}
z <- array(data=x,dim=c(3,3,3))
z
z[x]
z[rev(x)] # What does the function rev() do? How can you find out?
```

#### Browse the datasets available in R, find Edgar Anderson's Iris data and load it as a dataframe 

```{r}
data(iris)
head(iris)
dim(iris)
#view(iris)
```

#### Convert the iris dataset to a matrix and check the data types

```{r}
iris2 <- as.matrix(iris)
head(iris2)
mode(iris2)
```

```{r}
iris3 <- as.matrix(iris[,1:4])
head(iris3)
mode(iris3)
```

#### Using matrices instead of data frames 

When data is all numbers (except for row and column names), I normally use matrices rather than dataframes
If you have row names in your data file, and you want to convert to a matrix, you will need to specify which column has the row names (see the row.names argument in the read.table function)

# 3. Getting help with R
This is a summary of the information you can find here https://www.r-project.org/help.html

#### R Help: `help()` and `?`
The `help()` function and `?` help operator in R provide access to the documentation pages for **R functions, data sets, and other objects**, both for packages in the standard R distribution and for contributed packages. 
For example, to access documentation for the standard mean function, for example, enter the command:`help(mean)` or `help("mean")`, or `?mean` or `?"mean"` (the quotes are optional).

You may also use the `help()` function to access information about a **package** in your library - for example, `help(package="MASS")` - which displays an index of available help pages for the package along with some other information.

Help pages for functions usually include a section with executable examples illustrating how the functions work. You can execute these examples in the current R session via the `example()` 
```{r}
example(mean)
```

#### Searching for Help Within R

The `help()`  function and `?` operator are useful only if you already know the name of the function that you wish to use. There are also facilities in the standard R distribution for discovering functions and other objects. The following functions cast a progressively wider net. Use the help system to obtain complete documentation for these functions: for example, `apropos()`or `?apropos`.

The `apropos()` function searches for objects, including functions, directly accessible in the current R session that have names that include a specified character string. This may be a literal string or a regular expression to be used for pattern-matching. By default, string matching by `apropos()` is case-insensitive. For example, 

```{r}
apropos("^glm") 
```
returns the names of all accessible objects that start with the (case-insensitive) characters "glm"

#### R Help on the internet

There are internet search sites that are specialized for R searches, including `search.r-project.org` (which is the site used by RSiteSearch) and `Rseek.org`.

It is also possible to use a general search site like **Google**, by qualifying the search with "R" or the name of an R package (or both). 
**It can be particularly helpful to paste an error message into a search engine** to find out whether others have solved a problem that you encountered.

