---
title: "1.2.Graphing with R"
author: Adapted from Software Carpentry; modified by Maina,Stephanie
date: "27 October 2017"
output: html_document
## BASED ON: JOSH MADIN
## http://acropora.bio.mq.edu.au/resources/introduction-to-r/graphing/
##
##	MODIFIED BASED ON DELIVERY BY M. KOSNIK IN 2017.MAR

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Before starting**

**We need data to play with:**
  
```{r}
data <- read.csv("data/seed_root_herbivores.csv", as.is=TRUE)  # Read the csv file:

```
**What have we got here?**
```{r}
head(data)
```

  + Different ways reference columns, should all work

```{r results = "hide"}
data[,"Height"] 
data[["Height"]] 
data$Height
```

#######################################
### **1. Histograms**
#######################################
```{r eval=FALSE}
? hist()
```

#################
**Plot histogram of height**

```{r}
hist(data$Height)
```

**Lets add better axis labels**
```{r}
hist(data$Height, xlab='Height (cm)')
```

**Lets colour our columns & add main title**
```{r}
hist(data$Height, xlab='Height (cm)', col='blue', main="Figure 1")
```

**Another way to specify colours**
```{r}
hist(data$Height, xlab="Height (cm)", col=rgb(0,0.1,0.5), main="Figure 1")
```

#################
**Make a histogram for seed.heads... with no axes or labels**
```{r}
hist(data$Seed.heads, col='forestgreen', ann=FALSE, axes=FALSE)
```

**Add axes manually (allows for greater control)**

```{r eval=FALSE}
?axis()
```
      
```{r}
hist(data$Seed.heads, col='forestgreen', ann=FALSE, axes=FALSE)
axis(side= 1)	# identical to what would have been drawn from hist()
axis(2, lwd=2, lty=3, col='blue', font=3)	# specify details of line 
```

**Another way to add margin text (aka - label)**
  

```{r eval=FALSE}
?mtext()
```

```{r}
hist(data$Seed.heads, col='forestgreen', ann=FALSE, axes=FALSE)
mtext("Genus italicus", side=3, line=-2, font=4, adj=0.7, cex=2)	# big, italic font right side
mtext("Number of seed heads", side=1, line=2)
```

**What other options does histogram have?**

```{r eval=FALSE}
? hist()
```
 
  + **Set "breaks"**
```{r}
hist(sqrt(data$Seed.heads), breaks=(0:34), col='lightgrey')
```

**What else can "hist()" DO?**

```{r}
seedHeadHist <- hist(sqrt(data$Seed.heads), breaks=(0:36), plot=FALSE)
seedHeadHist
```

#######################################
### **2.  Bivariate plots**
#######################################

```{r eval=FALSE}
?plot()
```

**Specify the column with the x data and the column with the y data**

```{r}
plot(data$Height, data$Seed.heads)
```

**Alternatively, use formula notation (same plot, but better default axis labels)**
  
```{r}
plot(Seed.heads ~ Height, data)
```

**Plot it with a square root transform on the y axis?**
  
```{r}
plot(sqrt(Seed.heads) ~ Height, data) 
```

**Lets make nicer axis labels**

```{r}
plot(sqrt(Seed.heads) ~ Height, data, xlab="Height, cm", ylab="Sqrt number of seed heads", col='forestgreen')
```

**Lets make it plot lines instead**

```{r}
plot(sqrt(Seed.heads) ~ Height, data, xlab="Height (cm)", ylab="Sqrt number of seed heads", type = "l")
```

**Lets make it plot the axes, but not any points!**

```{r}
plot(sqrt(Seed.heads) ~ Height, data, xlab="Height, cm", ylab="Sqrt number of seed heads", type="n") 
```

**?? Why would we want to do that??**

```{r eval=FALSE}
?points()
```

**Lets plot the `"Root.herbivore == TRUE"` treatment as red circles**
  
```{r}
plot(sqrt(Seed.heads) ~ Height, data, xlab="Height, cm", ylab="Sqrt number of seed heads", type="n") 
points(sqrt(Seed.heads) ~ Height, data=data[data$Root.herbivore==TRUE,], col="red", pch=21) 
```

**Lets plot the `"Root.herbivore == FALSE"` treatment as green squares**
  
```{r}
plot(sqrt(Seed.heads) ~ Height, data, xlab="Height, cm", ylab="Sqrt number of seed heads", type="n") 
points(sqrt(Seed.heads) ~ Height, data=data[data$Root.herbivore==FALSE,], col="green", pch=22)
```  
  
**What else can we specify for points?**

Plot is super general (it will try to plot anything).. here it defaults to points, factors default to boxplots...

```{r}
plot(Seed.heads ~ as.factor(Seed.herbivore), data)
```

#######################################
### **3. Box plots**
#######################################

```{r eval=FALSE}
?boxplot()
```

```{r}
boxplot(data$Height, col='darkblue')
```

**SPLIT IT BY "Root.herbivore" TREATMENT?**

```{r}
boxplot(Height ~ Root.herbivore, data, col='skyblue') 
```

**Better axis labels**
```{r}
boxplot(Height ~ Root.herbivore, data, xlab="Root herbivores", ylab="Plant height, cm") 
```

**Add `"notch"` es?**

```{r}
boxplot(Height ~ Root.herbivore, data, xlab="Root herbivores", ylab="Plant height, cm", notch = TRUE)
```

**?? What are the "notches??**

Overlay the points? ( it works - if we make `root.herbivore` a factor )

```{r}
boxplot(Height ~ Root.herbivore, data, xlab="Root herbivores", ylab="Plant height, cm", notch = TRUE)
points(as.factor(data$Root.herbivore),data$Height, cex=0.5)
```


#######################################
### **4.Exploratory plotting**
#######################################

```{r eval=FALSE}
?pairs()
```

A very cool fucntion, but only works with numeric data.
A Great way to quickly access collinearity among variables.

```{r error=TRUE}
pairs(data)
```

**Which columns have numeric data??**
  
```{r}
names(data) 
comps <- names(data)[2:8]
comps
```

**Behold a quick set of plots for all pair-wise comparisons**
  
```{r}
pairs(data[comps])
```

#######################################
### **5. 3D plotting**
#######################################

**Lets use the built in volcano dataset for this...**

```{r eval=FALSE}
?volcano
```

**Two good built in functions are `image()` and `contour()`**
```{r eval=FALSE}
image(volcano) 
contour(volcano) 
```

**Specify the colour gradient**

```{r eval=FALSE}
image(volcano, col=terrain.colors(50))
```

**Add" contours over top of `image()`**

```{r eval=FALSE}
contour(volcano, add=TRUE)
```

#######################################
### **6.  Bar plots**
#######################################

I personally really dislike barplots, but they are commonly used.
R does not have a standard error function, but we can write one...

```{r}
standard.error <- function(x) { sd(x)/sqrt(length(x)) }
```

Remember all the way back to the morning...

USE `"aggregate"` TO GET THE mean `data['Height']` 

BY `data[c("Root.herbivore", "Seed.herbivore")]`

   + **TIP**: USING `data["Height"]` INSTEAD OF `data$Height` GIVES YOU NICER COLUMN NAMES

```{r}
mn <- aggregate(data["Height"], data[c("Root.herbivore", "Seed.herbivore")], mean) 
bp <- barplot(mn$Height)
```

** USE `"aggregate"` TO GET THE standard.error**
```{r eval=FALSE}
se <- aggregate(data["Height"], data[c("Root.herbivore", "Seed.herbivore")], standard.error)
```

** Add the standard error using arrows**
```{r eval=FALSE}
arrows(bp, mn$Height + se$Height, bp, mn$Height - se$Height, code=3, angle=90)
```

*Note: by assigning barplot() to bp... we can do the next step*

  
  ?? Doesn't quite look right

**Make a bar plot and specify y-axis "ylim" so that the whole se fits in...**

```{r eval=FALSE}
bp <- barplot(mn$Height, ylim=c(0, max(mn$Height + se$Height))) 
arrows(bp, mn$Height + se$Height, bp, mn$Height - se$Height, code=3, angle=90)
```

**Add labels**

```{r eval=FALSE}
axis(1, at=bp, labels=as.character(mn$Root.herbivore)) 
axis(1, at=bp, labels=as.character(mn$Seed.herbivore), line=1, tick=FALSE)  
```



? What does `tick` do?

**Make better labels**
  
```{r eval=FALSE}
herb_labels <- c("None", "Root only", "Seed only", "Both root & seed") 
bp <- barplot(mn$Height, ylim=c(0, 80)) 
arrows(bp, mn$Height + se$Height, bp, mn$Height - se$Height, code=3, angle=90)
axis(1, at=bp, labels=herb_labels)
```

Still not what we want because r does skips labels that would over write

```{r eval=FALSE}
axis(1, at=bp, labels=herb_labels, las=2)
title("Barplots are a pain...")
```

*How do we fix the margin size...*

*Show me how to do that crazy voodoo that youdoo?*

*Remember if you dig deep enough you can tinker with anything in r*

*Afraid? "You will be. You... will... be." (Yoda)* 


#######################################
##  **7. Global graphic parameters**
#######################################

```{r eval=FALSE}
?par
```

  + **NOTE**: No one remembers all of these parameters but hopefully you will remember that ?par gets you the answer
  + **NOTE**: Closing your graphing window will reset your global parameters (for good and bad)
  + **TIP**: You can save parameters to a variable to restore your old ones
  + **TIP**: par() Will tell you what the current parameters are

**Save current / original parameters**
  
```{r}
oldPar <- par()
```
So how can we fix our margin issue?
  
```{r eval=FALSE}
par(mar=c(9, 4, 4, 2)) 
bp <- barplot(mn$Height, ylim=c(0, 80)) 
arrows(bp, mn$Height + se$Height, bp, mn$Height - se$Height, code=3, angle=90, lwd=2)
axis(1, at=bp, labels=herb_labels, las=2) 
mtext("Type of herbivory", side=1, line=7) 
mtext("Plant height, cm", side=2, line=3) 
title("Barplots are still a pain...")
```

**Restore original parameters**
  
```{r error=FALSE}
par(oldPar)
```

#######################################
### **8. Saving plots**
#######################################

  + **NOTE**: You can save or "export" the graphing window to save your plots
  + **TIP**: Don't save or "export" the graphing window to save your plots
  + **TIP**: Always use a vector format (i.e., pdf) so that your plots look good when scaled
  + **TIP**: If you must use a raster format use png rather than jpeg
  + **TIP**: Save your plot from the command line!
        - Total control of plot sizes, etc to meet specific journal requirements
        - You can just run a script to regenerate all your plots if (when) your data change or your supervisor wants it to look a little different
        - Everytime the script runs you will always get the exact same plot, reproducibility is key to good science
  + **TIP**: Make a subdirectory called "figs" in your project directory
dir.create("./figs")
  + **NOTE**: Fancy script - only creates the directory if the directory is not already there
  + **TIP**: Dealing with the unexpected well is what makes flexible code (most code we write it not flexible)
if(!("./figs") %in% list.dirs(".")) dir.create("./figs")

**Start a pdf file to write to:**

```{r eval=FALSE}
par(mar=c(9, 4, 4, 2)) 
bp <- barplot(mn$Height, ylim=c(0, 80)) 
arrows(bp, mn$Height + se$Height, bp, mn$Height - se$Height, code=3, angle=90, lwd=2)
axis(1, at=bp, labels=herb_labels, las=2) 
mtext("Type of herbivory", side=1, line=7) 
mtext("Plant height, cm", side=2, line=3) 
title("Barplots are still a pain...")
pdf("./figs/my_barplot.pdf", width=5, height=5) 
```

What are the units of width and height?

Must always use `dev.off()` when you are done writing to the file... or it stays open...

```{r eval=FALSE}
par(mar=c(9, 4, 4, 2)) 
bp <- barplot(mn$Height, ylim=c(0, 80)) 
arrows(bp, mn$Heigh + se$Height, bp, mn$Heigh - se$Height, angle=90, code=3) 
axis(1, at=bp, labels=herb_labels, las=2) 
mtext("Degree of herbivory", 1, line=7) 
mtext("Plant height, cm", 2, line=3) 
title("Barplots are a pain...") 
dev.off()
```

**Start a png file**

```{r eval=FALSE}
png(filename="./figs/my_barplot2.png", width=400, height=400) 
par(mar=c(9, 4, 4, 2)) 
bp <- barplot(mn$Height, ylim=c(0, 80)) 
arrows(bp, mn$Heigh + se$Height, bp, mn$Heigh - se$Height, angle=90, code=3) 
axis(1, at=bp, labels=herb_labels, las=2) 
mtext("Degree of herbivory", 1, line=7) 
mtext("Plant height, cm", 2, line=3) 
title("Barplots are a pain...") 
dev.off()
```

  + **TIP**: When things go wrong saving plots
        - It is nearly always because the file was not closed
        - Repeat dev.off() until it says "cannot shut down device"


#######################################
### **9. Multiple panel plots**
#######################################

`"mfrow"` Fills your panels across the top row and then sequential rows down

`"mfcol"` Fills your panels down the left most column first...

```{r}
oldPar <- par()
```

**Make 4 histograms**

```{r eval=FALSE}
par(mfcol=c(2,2)) 
hist(data$Height, col='blue') 
hist(data$Height, col='green') 
hist(data$Height, col='red') 
hist(data$Height) 
par(oldPar)
```

**Lets make it nicer...**

    + Set margins all to 1
    
    + Set the outer margins to `c(4,3,4,2)`

```{r eval=FALSE}
par(mfrow=c(2, 2), mar=c(1, 1, 1, 1), oma=c(4, 3, 4, 2))
#lets only put the outer axes on (common if the axes are the same)
#lets label each panel a-d in the upper left corner
hist(data$Height, ann=FALSE, axes=FALSE, col='forestgreen') 
axis(2) 
mtext("A", 3, -2, adj = 0.1, font=2, cex=1.2) 
hist(data$Height, ann=FALSE, axes=FALSE, col='darkblue') 
mtext("B", 3, -2, adj = 0, font=2, cex=1.2) 
hist(data$Height, ann=FALSE, col='skyblue') 
mtext("C", 3, -2, adj = 0.1, font=2, cex=1.2) 
hist(data$Height, ann=FALSE, axes=FALSE, col='grey40') 
axis(1) 
mtext("D", 3, -2, adj = 0, font=2, cex=1.2)
```

Just to demonstrate the possible boxes...

```{r eval=FALSE}
?box()
box("plot", col="red") 
box("figure", col="blue") 
box("inner", col="black") 
box("outer", col="pink")
##	`mtext` really shines - putting joint axis labels  `"outer"`
mtext("Height (cm)", side=1, outer=TRUE, line=2) 
mtext("Frequency", side=2, outer=TRUE, line=2) 
mtext("Four plots of height", side=3, outer=TRUE, line=1, cex=1.5) 
```


#######################################
### **10.Adding additional things to plots**
#######################################

```{r eval=FALSE}
plot(sqrt(Seed.heads) ~ Height, data, type="p", axes=FALSE, ann=FALSE, pch=21, col="white", bg="black") 
axis(1) 
axis(2, las=2) 
mtext("Height (cm)", side = 1, line = 3) 
mtext("sqrt(Number of seed heads)", side = 2, line = 3) 
title("Figure 1", adj=0)

```

**Get/Plot a best fit line using `lm()`**

```{r }
mod <- lm(sqrt(Seed.heads) ~ Height, data)
```

**Plot the line using `abline`**

```{r eval=FALSE}
abline(mod, lwd=2, lty=2, col='blue')
```

**Get/plot the prediction intervals**

```{r eval=FALSE}
hs <- seq(min(data$Height), max(data$Height), 1) 
seed_pred <- predict(mod, list(Height = hs), interval = "prediction") 
lines(hs, seed_pred[,"lwr"], lty = 2) 
lines(hs, seed_pred[,"upr"], lty = 2)
```

**Plot a shaded confidence interval region**

```{r eval=FALSE}
polygon(c(hs, rev(hs)), c(seed_conf[,"lwr"], rev(seed_conf[,"upr"])), col = rgb(0,0,0,0.2), border = NA)
```

**Let's add the r-squared to the plot**

  + **TIP**: Unicode characters are an easy way to get some characters
```{r eval=FALSE}
text(80,5,paste("r\U00b2 = ",round(summary(mod)$r.squared,2))) # mod_sqrt does not exist: change to mod
```

**Add a legend**

```{r eval=FALSE}
legend("topleft", c("green data", "actual data", "orange point"), pch=c(4, 20, 3), col=c("green", "black", "orange"), bty="n")
```

Add some math 
```{r eval=FALSE}
text(90, 3, expression(Y[i] == beta[0]+beta[1]*X[i1]+epsilon[i])) # see demo(plotmath) for more examples 
```



Now... Wrap the super nice plots into a pdf...


```{r}
dev.off()
```



