# 3. Looking at your data: plotting

Plotting is covered in the next R module, but it's one of the best things about R so I can't resist showing how to do it:
  
  Here is a histogram of the height variable:
  
  ```{r}
hist(data$Height)
```
(it will appear in the bottom right of your screen)

Here is a scatter plot of Height vs weight:
  
  ```{r}
plot(data$Weight,data$Height)
```
The order of arguments is x-variable, y-variable.

There is an alternative interface using R's "formulae" - you'll see this a lot in statistical models. Read this as "`Height` is a function of `Weight`". It makes nicer axis labels, too.

```{r}
plot(Height ~ Weight, data)
```
Here is a series of bivariate plots for height, weight and the number of seed heads:
  
  ```{r}
pairs(data[c("Height","Weight","Seed.heads")])
```
The take-home being that R makes it very easy to create graphs, and most people who use it casually just make plots of whatever they're looking at. The plots can vary from quick and dirty like this to really beautiful pieces of art.


